---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# tidygm

Can we build and play music by manipulating tidy data frames ?

The {gm} package by Renfei Mao is young and promising, this package is a wrapper of {gm} where
I play with some ideas. You'll need to install {gm} first.

Install with

```r
remotes::install_github("moodymudskipper/tidygm")
```

I'll illustrate {tidygm} by recreating a version of Coldplay's song "Clocks".
I followed this great breakdown of the song by Rick Beato :

https://www.youtube.com/watch?v=7_warUFthJI&ab_channel=RickBeato

## 1st bar

```{r}
library(tidygm)
library(magrittr, include.only = "%>%")
```


The first bar of "Clocks" is a piano line that goes : 

> Eb6, Bb5, G5, Eb6, Bb5, G5, Eb6, Bb5

Let's make a song out of it using the function `new_song`, which creates a tidy data frame, and `play` which plays it. 

* The song is in F minor, the tempo is 129 bpm
* The meter is 4/4 (which is the default for `new_song`)
* We don't provide a `durations` argument so `new_song` will guess these are all eighth notes
* The default instrument is "piano" (at the moment it is the only instrument
supported by {gm} but this should hopefully change with the next minor release).

```{r}
piano_1 <- list("E-6", "B-5", "G5", "E-6", "B-5", "G5", "E-6", "B-5")
song <- new_song(list(piano_1), key = "Fm", tempo = 129)
song

play(song)
```

## 2nd bar

This worked well! let's write two bars in one go now :

```{r}
piano_2 <- list("C#6", "A#5", "F5", "C#6", "A#5", "F5", "C#6", "A#5")
song <- new_song(list(piano_1, piano_2), key = "Fm", tempo = 129)
song

play(song)
```

## 3rd bar

We don't want to build the object from scratch every time, so we provide utilities
to build our song data frame as we go.

The 3rd is the same as the 2nd, so we copy and paste it :

```{r}
song <- song %>% copy_and_paste(bar_id == 2)
song

play(song)
```

`copy_and_paste` has additional arguments `subset_to`, to indicate where to paste
(to overwrite values), and `what` to indicate what to paste, but if left empty
as done here we just copy everything to the end, with new values of `bar_id`.

## 4th bar

The 4th bar is different, since we cannot copy and paste it, we add
a new bar using `add_bars()`

```{r}
piano_4 <- list("C6", "A-5", "F5", "C6", "A-5", "F5", "C6", "A-5")
song <- add_bars(song, list(piano_4))
song

play(song)
```

## repeat it all

The full sequence is then repeated to complete the piano intro.

```{r}
song <- copy_and_paste(song, bar_id %in% 1:4)
song

play(song)
```

## and repeat it all again

The intro is then played again, but this time we'll add other instruments on top,
let's copy 4 additional bars only for now.

```{r}
song <- copy_and_paste(song, bar_id %in% 1:4)
song

play(song)
```


except instruments are added on top.

## add the bass

We add the bass at bars 9 to 12, we use the function `add_bars()` again
but with the `at` argument this time since we're not appending to the end.

The base line is very simple, we repeat for each bar an eighth note 8 times.

```{r}
bass_bars <- list(
  replicate(8, "D#3",  F),
  replicate(8, "A#2",  F), 
  replicate(8, "A#2",  F), 
  replicate(8, "F3" ,  F))
  
song <- add_bars(
  song, 
  pitches = bass_bars,
  instrument = "bass",
  at = 9:12)

song

play(song)
```

## add the guitar

On top of the bass a guitar is joining, the process is similar but now we show
how we deal with simultaneous notes (chords)

```{r}
guitar_bars <- list(
  replicate(8, c("G4","A#4", "D#5", "G5"), F),
  replicate(8, c("F4","A#4"), F),
  replicate(8, c("F4","A#4", "C#4"), F),
  replicate(8, c("G#4","C5", "D#5", "D#5"), F))
  
song <- add_bars(
  song, 
  pitches = guitar_bars,
  instrument = "guitar",
  at = 9:12)

song

play(song)
```

## finish the intro

Repeat the last 4 bars, with all instruments

```{r}
song <- copy_and_paste(song, bar_id %in% 9:12)
song

play(song)
```

## The end

We won't finish the song, but we have some building blocks we'd be able to reuse already.

{gm} has some limitations at the moment, and {tidygm} has more, but I see potential
in a tidy approach, it would be very easy for instance to filter out an instrument
from the mix here, or to shift the pitches up or down with appropriate functions.

I'm also interested in playing with modes and backward melodies to do automatically things like :

* https://www.youtube.com/watch?v=eGsthz78Ajo&ab_channel=Shred

* https://www.youtube.com/watch?v=gj4JnOZ2Ees&ab_channel=Shred

* https://www.youtube.com/watch?v=Sptr5yS5gAM&ab_channel=SteveCruickshank

I'm not sure if I'll move along with {tidygm} but thought I'd at least get the idea
out there.

